cILR: Taxonomic Enrichment Analysis with Isometric Log-Ratios

### 2021-06-02 tax_agg.tex

\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test. We added a pseudocount of 1 to all zero entry values. This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample. For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.

\noindent \textbf{Differential Abundance}: Since cILR are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on cILR generated scores. We added a pseudocount of 1 to all zero entry values. For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}.

\noindent \textbf{Disease Prediction}: We fit random forest on cILR scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section. We added a pseudocount of 1 to all zero entry values. Additionally, we also compared performance using enrichment scores against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.

@@

Panel A and B in figure \ref{fig:2} demonstrate type I error and power respectively across different simulation conditions. We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement. All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions. However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst. On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.

@@

These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}. Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations. As expected in Figure \ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods. Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05. Power analysis (Figure \ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.

To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set. This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated. Figure \ref{fig:3} presents this result. We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank sum test. Across both simulations (Figure \ref{fig:3}A) and real-data applications (Figure \ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios. In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2). Notably, the W-statistic provided the least information for classifying samples with inflated taxa.

@@

\subsection*{Differential abundance analysis}
Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank sum test) to perform differential abundance analysis. We compared the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
    \caption{Placeholder} 
    \label{fig:4}
\end{figure}

\subsection*{Disease Prediction} 
We can conceive of cILR as a feature engineering step before model fitting and prediction. Here we assessed the predictive performance of a standard machine learning model (random forest \cite{breiman2001}) using cILR scores compared to scores generated by other single sample enrichment methods in the gene set testing literature (ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}). Additionally, we also compared our results with the use of the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.

Figure \ref{fig:6} and \ref{fig:6} showed results from simulation studies as detailed in the \nameref{Methods} section under both classification and regression tasks. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio, however, they remained consistent across different levels of model saturation and inter-taxa correlation. Importantly, cILR methods outperformed both GSVA and ssGSEA, where the difference in AUC increases as a function of sparsity. Conversely, cILR scores did not perform as well as the traditional CLR approach, however, the performance gap decreased at higher sparsity levels. Similar patterns were observed for regression results (Figure \ref{fig:6}). Most interestingly, regression performance, in general, decreased at higher saturation levels (proportion of sets associated with the outcome). cILR remained more performant than both ssGSEA and GSVA, but was still less than the standard CLR, especially at higher signal-to-noise ratios. Similarly, cILR still performed much better in higher sparsity situations, where the results were almost identical to the CLR approach, especially in high saturation scenarios.

\begin{figure}[H]
    \centering
    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
    \caption{AUC scores of a random forest model for a binary outcome trained on cILR, ssGSEA, GSVA generated scores as well as on standard CLR transformed data evaluated on simulated data across sparsity levels, correlation, and signal-to-noise ratio. Panel \textbf{(A)} and \textbf{(B)} represent results across different levels of model saturation (proportion of sets associated with the outcome). cILR approaches outperformed GSVA and ssGSEA but not standard CLR.}
    \label{fig:6}
\end{figure}

@@

In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:7}. The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls. The Gevers et al. \cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al. \cite{nielsen2014} is a whole-genome shotgun sequencing data set. Similar to simulation experiments, across both data sets cILR methods provide much better performance than either GSVA or ssGSEA. Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set. Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
    \caption{Classification performance of a standard random forest model using cILR scores compared against existing methods in gene set testing literature and the standard centered-log ratio transformation approach. The learning task involves predicting patients with inflammatory bowel disease (including Crohn's disease and ulcerative colitis) versus controls. Data sets used span both 16S rRNA sequencing (Gevers et al. \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al. \cite{nielsen2014}). cILR marginally outperforms GSVA and ssGSEA in both data sets, but only performs better than CLR in the WGS data set.}
    \label{fig:7}
\end{figure}

@@

\subsection*{Inference with cILR}
We can use cILR to test for enrichment of taxa-sets at the sample level. Inter-taxa correlation can result in an inflated variance of resulting test statistics \cite{wu2012}, which results in an inflated type I error if not properly controlled for. Our method addresses this issue by combining the mean estimate of scores computed on permuted data and the variance estimate of scores computed on unpermuted data, where correlation is undisrupted. As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:2}). The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation. Conversely, the unadjusted cILR methods and the naive Wilcoxon rank sum test has inflated type I error even at low correlation levels ($\rho = 0.2$). However, the trade-off for proper type I error control is lower power. Simulation results showed that power is sensitive to sparsity across all correlation levels.

However, if there is no correlation between members of a set, cILR was still able to maintain 80\% power even with low effect size (fold change of 1.5), long as the set of interest is more densely populated, such as a set of core taxa that are prevalent across all samples. Unfortunately, this relationship doesn't hold for higher correlation levels. This means that in sets where associated taxa are highly correlated, a higher fold change is required for cILR to detect enrichment.

We also assessed cILR using different distribution forms, the normal distribution, and the mixture normal distribution. The difference in performance across the two choices was minimal, where both were able to control for type I error, but the normal distribution was slightly better when it comes to power only in certain simulation settings. This runs contrary to our distribution comparison analysis in Figure \ref{fig:1}, where the mixture distribution was the superior fit. We hypothesized that this might be due to the difficulty in fitting mixture distributions using the expectation-maximization algorithm (EM), as the convergence rate is slow when there is a lot of overlap between the mixtures and the mixing coefficients of one of the components are low \cite{naim2012}. Furthermore, there is a lot of degeneracy of parameter values in our variance adjustment procedure for the mixture distribution(Supplemental Material). These distribution fitting problems might introduce issues where the mixture distribution might not be the best fit. As such, improvements to mixture distribution fitting might allow for better performance to be extracted.

These results are replicated in our real data analysis, where we attempted to test for enrichment for a group of aerobic taxa among samples in the gingival site from HMP \cite{consortium2012}. Even though our labeling is not perfect, it is encouraging to see that the false discovery rate is low (approximately 0.05) when tested among a set of hypotheses where there is a mixture of true null and alternate situations. We believe that a practitioner can identify many interesting samples to follow up on by applying multiple testing adjustment procedures (such as Benjamini-Hochberg \cite{benjamini1995}) with cILR p-values and using less stringent criteria (e.g. 0.1 false discovery rate).

In short, we demonstrated that cILR can be used to test for significance testing at the sample level. Even though there is not a lot of difference between the distribution choices, ultimately cILR controls for type I error well despite tradeoffs in power. Even then, cILR will still be able to detect enriched sets in high correlation and sparsity situations insofar as the effect size is large. Researchers can utilize cILR to test for enrichment of a certain set of taxa of interest to perform follow-up experiments or to identify the prevalence of certain sub-population of taxa (for example, oxygen preference as in our real data example). The conservativeness of the test ensures confidence in detected samples.

\subsection*{Using cILR as inputs for downstream analyses}
Since cILR generates enrichment scores for each set at the sample level, they can be used in downstream analyses alongside standard statistical approaches. In this section we assessed the utility of cILR generated scores in two common analysis tasks in microbiome research: differential abundance testing and disease prediction.

For disease prediction, we benchmarked a basic random forest model \cite{breiman2001} to predict whether or not a patient has IBD given only their microbiome profile using cILR and relevant methods as inputs. Across both simulations and real data analyses, cILR scores perform better than alternative single sample scoring methods in the gene set testing literature (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}). This is consistent with previous analysis (Figure \ref{fig:3}), where we assessed the informativeness of cILR generated scores by using AUC to measure whether samples with high scores correspond with the true label of an inflated set. Results indicated that cILR scores are more adaptive to microbiome data sets than GSVA and ssGSEA. However, cILR scores underperform the standard approach of applying the CLR transformation to count aggregated data, except for the WGS real data analysis. Despite this, there are still benefits to using cILR scores in the context of downstream modeling beyond random forest predictive models. First, CLR transformed variables have a sum to zero constraints, making the covariance matrix is singular \cite{gloor2017}, impacting methods that rely on matrix decomposition such as principal component regression or canonical correlation analysis. Second, cILR provides an additional value of significance testing for enrichment for sample screening. Third, CLR transformed variables still involve the count summation procedure for variable aggregation, therefore comparison between sets of different sizes remains unviable.

Despite such drawbacks, cILR methods still produce competitive performance values across both regression and classification tasks even in low signal-to-noise settings. Most interestingly, the performance gap between cILR and CLR was smaller in high sparsity settings ($p = 0.6$), which is a feature of microbiome relative abundance data \cite{li2019}. Furthermore, these predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), and good performance also indicated that cILR scores can also capture the joint distribution of considers sets, especially with regards to their relationship with a defined outcome such as IBD status.

Unsurprisingly, different variants of cILR did not differ significantly in performance similar to significance testing experiments. Even though we expect Z-scores to perform better due to the added information of directionality of enrichment, using CDF values did not alter performance in both the single sample AUC evaluations (Figure \ref{fig:6}) as well as the random forest evaluations (Figure \ref{fig:6}).

In short, users can apply cILR generated enrichment scores as inputs to downstream analyses, especially predictive models and differential abundance analysis. cILR scores are informative in discriminating samples between classes even in low effect size situations. Even though cILR did not outperform a standard analysis of count aggregation followed by the CLR transformation, the additional benefits listed above would prove to be attractive in certain use cases, especially in high sparsity situations.

\subsection*{Limitations and future directions}
There are various limitations to using cILR to generate sample-level taxonomic aggregation for microbiome data. First, significance testing at the sample level is not powered to assess inflated sets with low effect size. Practitioners can be confident in the validity of a significant result (following proper multiple testing correction) due to cILR's type I error control, but should not expect cILR to be able to identify all samples with inflated counts. Second, even though cILR is flexible and addresses the compositional nature of microbiome data, the performance of cILR might not be consistently better than competing methods. Careful evaluation of methods should be done before application to ensure that the value of cILR scores is appropriately extracted in suitable models. Third, cILR did not directly address the zero-inflatedness of microbiome data and utilizes a pseudocount to ensure log operations are valid. As such, practitioners are encouraged to choose a different pseudocount or an appropriate method of choice.

As such, we hope to address the aforementioned issues with cILR in later projects. First, we hope to refine the distribution fitting procedure for the mixture distribution due to its superior fit and address the identifiability problem that might underlie the lack of power in our experiments. Second, we hope to incorporate some of the sophisticated model-based zero correction methods \cite{martin-fernandez2012,kaul2017a}. Third, we can account for the difference in importance of certain taxa by incorporating taxa-specific weights. Finally, we also hope to curate more interest apriori sets that are functionally informative, which would generate more interesting insight and improve interpretability compared to using taxonomic categories such as Phylum or Genus.

\section*{Conclusion}
Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomic data sets. However, there has not been a lot of set-based analysis methods developed specifically for microbiome relative abundance data. In this manuscript, we introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level. We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.

>>>>

\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test. We added a pseudocount of 1 to all zero entry values. This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample. For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test. All three approaches were applied directly on count data (after pseudocount). For GSVA, the Poisson kernel was used.

\noindent \textbf{Differential Abundance}: Since cILR are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on cILR generated scores. We added a pseudocount of 1 to all zero entry values. For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}. For DESeq2, we performed a likelihood ratio test against an intercept only reduced model with dispersion estimated with local fit. For corncob, we also performed a likelihood ratio test against an intercept only reduced model without bootstrapping.

\noindent \textbf{Disease Prediction}: We fit random forest on cILR scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section. We added a pseudocount of 1 to all zero entry values. Additionally, we also compared performance using cILR against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.

@@

\subsubsection*{Simulation studies}
Panel A and B in figure \ref{fig:2} demonstrate type I error and power respectively across different simulation conditions. We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement. All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions. However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst. On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.

@@

To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set. This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated. Figure \ref{fig:3} presents this result. We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $W$ test statistic from the Wilcoxon rank sum test. Across both simulations (Figure \ref{fig:3}A) and real-data applications (Figure \ref{fig:3}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios. In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2). Notably, the W-statistic provided the least information for classifying samples with inflated taxa.

\subsubsection*{Real data evaluations}  
These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}. Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations. As expected in Figure \ref{fig:2}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods. Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05. Power analysis (Figure \ref{fig:2}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.

@@

\subsection*{Differential abundance analysis}
cILR generates sample-specific scores representing the degree of enrichment of a pre-defined set. As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test). We compared the performance of this approach with cILR and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.

\subsubsection*{Simulation studies}
\subsubsection*{Real data evaluations}
We also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP. For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different. For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.   
We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of cILR were controlling for type I error at the defined $\alpha$ threshold of 0.05. Conversely, true positive rate for all methods were roughly more similar. As expected, using the Wilcoxon rank sum test demonstrated have power true positive rate, but the difference was not large. Surprisingly, using Welch's t-test demonstrated similar true positive rate as both corncob and DESeq2, while also controlling for type I error. Choice of underlying distribution for cILR did not seem to alter performance.

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
    \caption{Differential abundance analysis using corncob, DESeq2 and cILR with either Wilcoxon rank sum test or Welch's t-test. Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample. Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes. Both evaluations use 16S rRNA gene sequencing data from HMP. Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site. Results showed that cILR associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2} 
    \label{fig:4}
\end{figure}

\subsection*{Disease Prediction}   
Since cILR can generate informative scores that can discriminate between samples with inflated counts for a set (Fig. \ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models. In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating cILR, ssGSEA, and GSVA). Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxa sets aggregated via abundance summations.

\subsubsection*{Simulation studies}
Fig. \ref{fig:6} showed results for simulation studies as detailed in the \nameref{Methods} section. Panel A. presents results for a regression learning task with a continuous outcome while panel B. presents results for a classification task with a binary outcome. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio. Both CLR and cILR approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks. This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while cILR and CLR did not experience a similar level of impact. As such, performance gap widens with increasing correlation and sparsity. Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.

Unfortunately, cILR did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}. This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation. However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR. Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity. Additionally, we did not observe any performance difference between the different variations of cILR.

\begin{figure}[H]
    \centering
    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio. Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure. Panel \textbf{(B)} presents performance on a classification task with AUC as the evaluation measure. cILR approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
    \label{fig:6}
\end{figure}

@@

\subsubsection*{Real data evaluations}
In addition to parametric simulations, we also assessed the performance of using cILR scores in predictive models with real data sets. Fig. \ref{fig:7} presents results for two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition. The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al. \cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.6]{figures/data_prediction_plot.png}
    \caption{Predictive performance of a naive random forest model trained on cILR, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles. Data sets used span both 16S rRNA gene sequencing (Gevers et al. \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al. \cite{nielsen2014}). cILR performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
    \label{fig:7}
\end{figure}

Similar to simulation experiments, we also fitted a naive random forest model using CILR, ssGSEA, GSVA, or CLR transformed variables as inputs, and use AUC as the performance criteria. Results also replicated that of the simulations, where across both data sets cILR and CLR methods provide much better performance than both GSVA or ssGSEA. Interestingly, the cILR approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set. However, these results indicate that cILR generated scores are informative, providing competitive performance when acting as inputs to disease predictive models. Most importantly, performance values are consistent across both simulated and real data sets.

@@

\subsection*{Inference with cILR}
The formulation of cILR as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}. This allows for conducting inference with cILR even at the sample-level. We assessed the usage of cILR in this type of analysis by evaluating type I error and power across both simulation studies and real data applications. Most importantly, we demonstrated that our adjusted cILR approach was able to address the issue of variance inflation due to correlation \cite{wu2012} by controlling for type I error at the appropriate $\alpha$ level across different levels of simulated inter-taxa correlation (Fig. \ref{fig:2}) while conversely unadjusted cILR and the naive Wilcoxon rank sum test showed much higher rates of error. This is further encouraged in real data analysis where the false discovery rate was around $0.05$ when a collection of true null and true alternate hypotheses were tested. Unfortunately the trade-off between good type I error control is lower power. The conservativeness of the test attenuates with higher sparsity and correlation, where power was not approaching even 50\% even at the highest effect sizes. However, when the degree of such data features are reasonable, cILR will still be able to detect a reasonable proportion of samples with inflated counts.

We also observed that choosing different distributional forms did not alter performance values for cILR. This runs contrary to our comparison analysis in Fig. \ref{fig:1}, where we demonstrated that the mixture normal distribution had superior fit compared to the simple normal for raw cILR scores computed under the global null. We hypothesized that this might be due to the difficulty in fitting mixture distribution to data using the expectation maximization algorithm, as the convergence rate is slow when there is high overlap between the mixtures, resulting in a small mixing coefficient for one of the components \cite{naim2012}. As such, in our implementation of cILR, in order to ensure convergence for the estimating procedure we increased the number of iterations while relaxing the tolerance parameter. Furthermore, there are also possible problems with our adjustment procedure for the mixture distribution that might impact overall fit. In order to combine the scale and location estimate of two mixture distributions, we fixed the overall mean, standard deviation, mixing coefficient and component-wise means and used an optimization procedure to find the component-wise variances. However, this means that we have one equation for the overall variance but two possible parameters to estimate. As such, there is no guaranteed unique solution to component-wise variances. We hypothesized that the instability and degeneracy in component-wise variance estimates might impact the fidelity of estimates at the tails of the distribution, thereby affecting inference.

Despite these concerns, empirical results still indicate that cILR can confidently identify samples with inflated counts. The conservativeness of the correlation adjustment procedure ensures that significant results can be trusted by practitioners, even if cILR might not be able to exhaustively identify all samples with inflated counts. In situations where either the data is less sparse (e.g. containing a lot of core taxa that are prevalent across all samples), there is less inter-taxa correlation within the set (e.g. taxa that do not participate in common pathways but have shared characteristics like pathogenicity), or if the effect size is large, then cILR will still be able to produce reasonable power. A practitioner can use cILR to screen for samples for subsequent analysis that might involve significant costs, or perform hypothesis generation using a less stringent criteria alongside a multiple testing adjustment procedure (such as Benjamini-Hochberg \cite{benjamini1995}).

\subsection*{Downstream analysis}
The sample-level enrichment scores generated by the cILR method can be used in downstream analyses commonly performed in microbiome research: differential abundance testing and disease prediction.

\subsubsection*{Differential abundance analysis}
For differential abundance testing, we evaluated whether using cILR alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes. We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014}. We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome data \cite{mcmurdie2014}. Conversely, corncob is a newer method developed specifically for microbiome data sets, where taxonomic counts are modeled directly newly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2. In our analyses, we found that both corncob and DESeq2 has inflated levels of type I error.

\subsubsection*{Predictive models}
For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs. Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA. Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed. This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}. Results indicated that cILR produce good performance values across both real data analysis and simulation scenarios. Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicate that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses. Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome data sets. This is consistent with our sample ranking analysis (Fig. \ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts. However, cILR did not outperform the CLR approach across all our simulation studies, and only marginally performed better in the real data analysis with WGS data.

However, in simulation studies, this performance gap between CLR and cILR decreases with higher sparsity and correlation, especially in low effect saturation scenarios. Additionally, there are also downsides to applying CLR. First, the singular covariance matrix of CLR transformed variables is singular due to a sum to zero constraint \cite{gloor2017}, preventing the proper usage of approaches that rely on matrix decomposition. Second, the procedure still relies on using summation of counts prior to transformation, which means that we still can't compare across sets of different sizes, and any bias might still be propagated \cite{mclaren2019}. As such, despite benefits in performance for a naive random forest model, there is still space for using cILR as primary inputs into predictive models.

Similar to other experiments in downstream usage of cILR, performance did not change with different underlying distributions, output types, or correlation status. This is surprising since we expect z-scores to perform better as they are able to capture the direction of an association. The fact that this effect persisted even onto our real data analysis suggests that this is not due to a deficiency of our simulation design. As such, practitioners who wish to use cILR in predictive models might be suited to use the settings that is the fastest to compute.

Ultimately, results indicate that cILR can produce informative scores that contribute to competitive performance of prediction models even in low signal-to-noise ratios with high inter-taxa correlation and sparsity. Even though there exists situations where it might not provide maximum predictive values, the flexibility of cILR in various types of analyses enable ven though in some scenarios it might not provide maximum predictive values.

\subsection*{Limitations and future directions} 
There are various limitations to our evaluation of cILR. First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome data. There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation section of prior studies. Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is flawed. This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut. As such, results from power analyses using this data set is only relative between the comparison methods, and absolute values should should not carry meaning. Finally, we assumed that taxa within a set all be equally associated with the outcome. This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.

However, our evaluation still showed various drawbacks of the cILR method. First, inference with cILR is limited in being able to exhaustively detect all samples with significant inflated counts for a set in situations where there is a high degree of sparsity and inter-taxa correlation. Second, for downstream analyses, cILR might not always better than competing methods, especially when being used as inputs to predictive models. We hypothesized that this might be due to the lack of fit for the underlying null distribution in high correlation settings, especially the identifiability problem associated with adjusting the mixture normal distribution. As such, we hope to refine the null distribution estimating procedure by either choosing a better distributional form, or to further constrain the optimization procedure of the mixture normal distribution by fixing the third and fourth moments.

In addition, there are possible extensions cILR can we can consider to provide more flexibility across different types of scenarios in data analysis. First, cILR did not address the sparsity of microbiome data and relies on pseudocount to ensure log operations are valid. We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}. Second, cILR also treated all taxa within the set as equally contributing to the set. Incorporate taxa-specific weights can reduce the influence of outliers, such as rare or highly invariant taxa. Finally, curating sets based on apriori characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.

\section*{Conclusion}
Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets. However, there has not been a lot of set-based analysis methods developed specifically for microbiome relative abundance data. In this manuscript, we introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level. We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.
